\thornado-hydro: towards discontinuous galerkin methods for supernova hydrodynamics

### 2018-10-16 Documents/Euler/Astronum_2018/ms_endeve.tex
Edits due to comments for Sam.  Also edited the reference list

\begin{document}
\title{\thornado-hydro: towards discontinuous galerkin methods for supernova hydrodynamics\footnote{This manuscript has been authored by UT-Battelle, LLC under Contract No. DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government retains and the publisher, by accepting the article for publication, acknowledges that the United States Government retains a non-exclusive, paid-up, irrevocable, world-wide license to publish or reproduce the published form of this manuscript, or allow others to do so, for United States Government purposes. The Department of Energy will provide public access to these results of federally sponsored research in accordance with the DOE Public Access Plan(http://energy.gov/downloads/doe-public-access-plan).}}

\author{Eirik Endeve$^{1,2,6}$, Jesse Buffaloe$^{2}$, Samuel Dunham$^{3}$, Nick Roberts$^{2}$, Kristopher Andrew$^{4,1}$, Brandon Barker$^{2}$, David Pochik$^{2}$, Juliana Pulsinelli$^{5,7}$ and Anthony Mezzacappa$^{2,6}$}
\address{$^{1}$Computer Science and Mathematics Division, Oak Ridge National Laboratory, TN 37831}
\address{$^{2}$Department of Physics and Astronomy, University of Tennessee Knoxville, TN 37996}
\address{$^{3}$Department of Astronomy, Vanderbilt University, TN 37212}
\address{$^{4}$Department of Physics and Astronomy, University of Kentucky Lexington, KY 40506}
\address{$^{5}$Department of Astrophysical Sciences, Princeton University, NJ 08544}
\address{$^{6}$Joint Institute for Computational Sciences, Oak Ridge National Laboratory, TN 37831}
\address{$^{7}$Physics Division, Oak Ridge National Laboratory, TN 37831}
\ead{endevee@ornl.gov}

\begin{abstract}
The toolkit for high-order neutrino-radiation hydrodynamics (\thornado) is being developed for simulations of core-collapse supernovae (CCSNe) and related problems.  
Current capabilities in \thornado\ include solvers for the Euler equations --- in non-relativistic and special relativistic limits --- and the two-moment model of neutrino transport.  
The spatial discretization in \thornado\ is based on the discontinuous Galerkin (DG) method, which is receiving increased attention from the computational astrophysics community.  
In this paper, we provide an overview of the numerical methods for the Euler equations in \thornado, and present some encouraging preliminary numerical results from a set of basic tests in one and two spatial dimensions.  
\end{abstract}

@@

The toolkit for high-order neutrino-radiation hydrodynamics (\thornado) is being developed to simulate neutrino-radiation hydrodynamics in CCSNe and related applications in nuclear astrophysics.  
The spatial discretization of solvers for hyperbolic partial differential equations in \thornado\ is based on the DG method, while we foresee using a combination of spectral and continuous finite element methods for solving elliptic equations; e.g., for Newtonian gravity or relativistic gravity employing the conformal flatness approximation \cite{wilson_etal_1996}.  
With these approaches, we employ high-order discretization techniques based on a common mathematical framework in all the major model components.  
In this paper, we provide an initial description and encouraging demonstration of the DG method implemented in \thornado\ to solve the Euler equations.  
We note that others (e.g., \cite{radiceRezzolla_2011,schaal_etal_2015}) have also investigated the DG method in the context of astrophysical hydrodynamics and reported encouraging results.  
Whether the high-order approach will improve accuracy and efficiency of CCSN models remains to be demonstrated.

\section{Discontinuous Galerkin Method}

@@

We let the approximation space for the DG method, $\mathbb{V}^{k}$, be constructed from the tensor product of one-dimensional polynomials of maximal degree $k$.  
Note that functions in $\mathbb{V}^{k}$ can be discontinuous across element interfaces.  
The semi-discrete DG problem is to find $\vect{U}_{h}\in\mathbb{V}^{k}$, which approximates $\vect{U}$ in Eq.~\eqref{eq:extendedEulerCompact}, such that for all $v\in\mathbb{V}^{k}$ and all $\bK\in\mathcal{T}$
\begin{align}
  &\partial_{t}\int_{\bK}\vect{U}_{h}\,v\,dV
  +\sum_{i=1}^{d}\int_{\partial\bK^{i}}\big(\,\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\vect{U}_{h})\,v\big|_{\xH^{i}}-\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\vect{U}_{h})\,v\big|_{\xL^{i}}\,\big)\,d\tilde{\bx}^{i} \nonumber \\
  &\hspace{24pt}
  -\sum_{i=1}^{d}\int_{\bK}\vect{F}^{i}(\vect{U}_{h})\,\pd{v}{i}\,dV
  =\int_{\bK}\vect{S}(\vect{U}_{h})\,v\,dV.  
  \label{eq:semidiscreteDG}
\end{align}
In Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\vect{F}}^{i}(\vect{U}_{h})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
The numerical flux function $\vect{f}^{i}$ is evaluated using values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\vect{F}}^{i}(\vect{U}_{h})=\vect{f}^{i}(\vect{U}_{h}(x^{i,-},\tilde{\bx}^{i}),\vect{U}_{h}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g., in the arguments of $\vect{U}_{h}$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
We use the Harten-Lax-van Leer (HLL) flux \cite{harten_etal_1983} or the HLLC flux \cite{toro_etal_1994,mignoneBodo_2005} for all the numerical experiments presented in Section~\ref{sec:numerical}.

We provide further details on the DG method to arrive at the equations that are actually evolved in \thornado.  
We start by introducing some notation, defining the polynomial expansion for $\vect{U}_{h}$ and the quadrature rules used to evaluate the integrals in Equation~\eqref{eq:semidiscreteDG}.  
Then we provide explicit expressions for each of the terms in Equation~\eqref{eq:semidiscreteDG}.  
For simplicity we consider one spatial dimension ($d=1$) and drop indices denoting the spatial dimension.  
In each element $K$, we use a nodal representation of the conserved variables $\vect{U}$; i.e.,
\begin{equation}
  \vect{U}(x,t)\approx
  \vect{U}_{h}(x,t)=\sum_{i=1}^{N}\vect{U}_{i}(t)\,\ell_{i}(x),
  \quad\text{where}\quad
  \ell_{i}(\eta)=
  \prod_{\substack{j=1\\j\ne i}}^{N}\f{\eta-\eta_{j}}{\eta_{i}-\eta_{j}}
  \label{eq:conservedNodalExpansion}
\end{equation}
are Lagrange polynomials defined on the reference element $I = \{ \eta : \eta \in (-0.5,0.5) \}$, and constructed to interpolate the node set $S_{N}=\{\eta_{i}\}_{i=1}^{N}\subset I$.  
The spatial coordinate $x$ and the reference coordinate $\eta$ are related by the mapping $x(\eta)=\xL+(0.5+\eta)\,\dx$.  
Then, for any $\eta_{j}\in S_{N}$, $\ell_{i}(\eta_{j})=\delta_{ij}$, so that $\vect{U}_{h}(x(\eta_{j}),t)=\vect{U}_{j}(t)$.  
We introduce numerical quadratures to evaluate the integrals in Equation~\eqref{eq:semidiscreteDG}.  
First we define the $M$-point quadrature $Q_{M}:C^{0}(I)\to\mathbb{R}$ with abscissas $\hat{S}_{M}=\{\eta_{q}\}_{q=1}^{M}$ and weights $\{w_{q}\}_{q=1}^{M}$, normalized such that $\sum_{q=1}^{M}w_{q}=1$.  
The $M$-point Legendre-Gauss quadrature, which we use, integrates polynomials of degree $\le 2M-1$ exactly.  
Then, if $P_{h}(x)$ is such a polynomial, we have
\begin{equation}
  \f{1}{\dx}\int_{K}P_{h}(x)\,dx=\int_{I}P_{h}(\eta)\,d\eta=Q_{M}\big[P_{h}\big]\equiv\sum_{q=1}^{M}w_{q}\,P_{h}(\eta_{q}).  
\end{equation}

@@

\section{Preliminary Numerical Results}
\label{sec:numerical}

In this section we show preliminary numerical results obtained with the DG method implemented in \thornado\ to solve the non-relativistic and special relativistic Euler equations.  
We use the ideal gas equation of state where the pressure $p$ is related to the internal energy density $e$ by $p=(\Gamma-1)e$, and where $\Gamma$ is the (constant) adiabatic index.  
Let the $d$-dimensional computational domain be $D=\times_{i=1}^{d}[\xMin^{i},\xMax^{i}]$, where $\xMin^{i}$ and $\xMax^{i}$ denote the coordinates of the inner and outer boundary in the $i$-th coordinate direction, respectively.  
We use polynomials of degree $k=2$ combined with SSP-RK3 time stepping, and all the tests were run with a Courant-Friedrichs-Lewy (CFL) factor of $C=0.1$.  
We use the HLL Riemann solver when solving the non-relativistic Euler equations and the HLLC Riemann solver when solving the special relativistic Euler equations.  
The purpose of the tests is gauge the performance of the DG implementation on a set of benchmarks as an initial measure of its suitability for future for CCSN simulations.

@@

For the NR Euler equations, the state, flux, and source vectors in Eq.~\eqref{eq:extendedEulerCompact} are
\begin{equation}
  \vect{U}=\big(\rho,\rho u_{j},E\big)^{\trans},~
  \vect{F}^{i}=\big(\rho\,u^{i},\Pi^{i}_{~j},(E+p)u^{i}\big)^{\trans},~
  \text{and}~
  \vect{S}=\big(0,\f{1}{2}\Pi^{ik}\pd{\gamma_{ik}}{j}-\rho\pd{\Phi}{j},-\rho u^{i}\pd{\Phi}{i}\big)^{\trans},
\end{equation}
where $\rho$ and $u^{i}$ are the mass density and components of the fluid three-velocity, respectively.  
The stress tensor is $\Pi^{i}_{~j}=\rho u^{i} u_{j}+p\delta^{i}_{~j}$, and $E=e+\f{1}{2}\rho u_{i} u^{i}$ is the fluid (internal plus kinetic) energy density.  
The sources are due to curvilinear coordinates and Newtonian gravity.

\subsubsection{Sod Shock Tube}

@@

\subsubsection{Sedov-Taylor Blast Wave}

@@

\subsubsection{Shu-Osher Shock Tube}

@@

\subsubsection{Kelvin-Helmholtz (KH) Instability}
This hydrodynamic instability is of significant astrophysical relevance --- including in the explosion phase of CCSNe --- and may occur in the interface separating fluids in relative motion.  
We compute this problem on the unit square $D=[0,1]\times[0,1]$, imposing periodic boundary conditions, and with initial conditions and single-mode perturbation taken from \cite{mcnally_etal_2012}.  
We use $256^{2}$ elements and evolve until $t=3.0$, well into the nonlinear regime.  
Results are displayed in Figure~\ref{fig:KelvinHelmholtz}.  
We have computed two models, one with $C_{\TCI}=0.06$ (left panels) and one without limiting ($C_{\TCI}\to\infty$; right panels).

@@

Early on ($t=1.5$), the results from the two runs are practically indistinguishable, show no sign of developing secondary billows, and agree visually with results presented in \cite{mcnally_etal_2012}.  
When $t=2.4$, the run with $C_{\TCI}=0.03$ has started to develop secondary billows within the coiled-up interface separating dense (white) and light (black) fluids, while the model with no limiting has not.  
When $t=3.0$, these secondary instabilities appear to have largely destroyed the coil structure for the model with limiting, while the coils remain intact in the model without limiting (although secondary billows have started to form in this model at this time).  
As discussed in more detail by \cite{mcnally_etal_2012}, the development of the secondary billows may be an artifact of numerical perturbations and diffusion, which is delayed when the resolution is increased.  
Our results are consistent with this in the sense that the less diffusive model develops secondary billows later.

\subsubsection{Liska-Wendroff Implosion}

@@

\subsubsection{Standing Accretion Shock (SAS)}

@@

\subsubsection{Yahil-Lattimer Collapse}

The final test of the NR hydrodynamics in \thornado\ involves self-gravity and is due to \cite{yahilLattimer_1982,yahil_1983}.  
It models the self-similar collapse of a polytropic star; i.e., $p=\kappa\rho^{\Gamma}$, where $\kappa$ is the polytropic constant.  
This test is also of immediate relevance to CCSN simulations.  
In \cite{yahilLattimer_1982,yahil_1983}, self-similar solutions to the gravitational collapse problem where constructed for $6/5\le\Gamma<4/3$.   
The solutions smoothly connect an homologously collapsing subsonic inner core (velocity proportional to radius) to a supersonically collapsing outer core in near free-fall ($u\propto r^{-1/2}$).  
\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/YahilLattimerCollapse_MassDensity_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/YahilLattimerCollapse_Velocity_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:YahilLattimer}Results from the Yahil-Lattimer collapse test using $256$ elements.  Results obtained with \thornado\ (dashed black) are compared with the reference solution from Yahil \cite{yahil_1983} (solid red).  The mass density (left panel) and velocity (right panel) are plotted versus radius.  We compare the solutions at select central densities during collapse, approximately $[10^{10},10^{11},10^{12},10^{13},10^{14}]$~g~cm$^{-3}$, which correspond to $(-t)=[51.0,15.0,5.0,1.5,0.5]$~ms.  }
\end{figure}
With two dimensional parameters in the model (the gravitational constant $G$ and the polytropic constant $\kappa$), the dimensionless similarity variable is
\begin{equation}
  X = \kappa^{-1/2} \, G^{(\Gamma-1)/2} \, r \, (-t)^{\Gamma-2},
\end{equation}
where the origin of time is the moment of infinite central density.  
All the hydrodynamic variables can be expressed as a function of $X$, and the time-dependent Euler equations can be recast as a system of ODEs (see \cite{yahil_1983} for details).  
To construct a reference to compare with our numerical results we have solved the ODEs given in \cite{yahil_1983} to obtain these self-similar solutions.

We show results for a model with $\Gamma=1.3$.  
We use spherical polar coordinates, and let $D=[0,1\times10^{5}]$~km, which is covered with $256$ elements.  
The Newtonian gravitational potential is obtained by solving Poisson's equation using a third-order accurate continuous finite element method (e.g., \cite{brennerScott_2008}).  
We use a geometric grid to resolve the mass distribution as the star collapses and the central density increases from about $10^{9}$~g~cm$^{-3}$ to about $10^{14}$~g~cm$^{-3}$.  
The size of the innermost element is set to $1$~km while the last element is about $3\times10^{3}$~km.  
We specify the polytropic constant by setting a reference pressure $p=6\times10^{27}$~erg~cm$^{-3}$ for $\rho=7\times10^{9}$~g~cm$^{-3}$ (reasonable values for a massive star in the pre-collapse stage).  
We also specify the collapse time to $(-t)=150$~ms.  
Results comparing the mass density and velocity of Yahil's reference solution to the solution obtained with \thornado\ using $C_{\TCI}=0.03$ are plotted in Figure~\ref{fig:YahilLattimer}.  
The agreement of the mass density and velocity profiles is good throughout collapse.

@@

\section{Summary and Outlook}

We have presented preliminary algorithm details and numerical results for solvers of the non-relativistic and special relativistic Euler equations of gas dynamics as implemented in the toolkit for high-order neutrino-radiation hydrodynamics (\thornado).  
The spatial discretization is based on the DG method and the ODEs resulting from this discretization are integrated in time using SSP-RK methods.  
We employ a spectral-type nodal collocation DG approximation based on Legendre-Gauss points for interpolation and numerical quadrature evaluation \cite{bassi_etal_2013}.  
This choice simplifies expressions for the semi-discretized equations, especially when applied to problems involving curvilinear coordinates encoded in a metric (e.g., in numerical relativity).  
Results from a suite of tests in one and two spatial dimensions --- including problems with strong shocks --- demonstrate reliable performance of the implementation.  
Two tests in spherical symmetry (standing accretion shock \cite{blondin_etal_2003} and Yahil-Lattimer collapse \cite{yahilLattimer_1982,yahil_1983}) demonstrate the DG method's ability to handle conditions relevant to simulations of CCSN.  
We will present a more in-depth analysis of the performance of \thornado\ on these (and related) problems in a future study.

@@

\ack E.~E. and A.~M. acknowledge support from the NSF Gravitational Physics Program (NSF-GP 1505933 and 1806692).

>>>>

\begin{document}
\title{\thornado-hydro: towards discontinuous galerkin methods for supernova hydrodynamics\footnote{This manuscript has been authored by UT-Battelle, LLC under Contract No. DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government retains and the publisher, by accepting the article for publication, acknowledges that the United States Government retains a non-exclusive, paid-up, irrevocable, world-wide license to publish or reproduce the published form of this manuscript, or allow others to do so, for United States Government purposes. The Department of Energy will provide public access to these results of federally sponsored research in accordance with the DOE Public Access Plan (http://energy.gov/downloads/doe-public-access-plan).}}

\author{Eirik Endeve$^{1,2,6}$, Jesse Buffaloe$^{2}$, Samuel J Dunham$^{3}$, \\ Nick Roberts$^{2}$, Kristopher Andrew$^{4,1}$, Brandon Barker$^{2}$, \\ David Pochik$^{2}$, Juliana Pulsinelli$^{5,7}$ and Anthony Mezzacappa$^{2,6}$}
\address{$^{1}$Computer Science and Mathematics Division, Oak Ridge National Laboratory, TN 37831}
\address{$^{2}$Department of Physics and Astronomy, University of Tennessee Knoxville, TN 37996}
\address{$^{3}$Department of Astronomy, Vanderbilt University, TN 37212}
\address{$^{4}$Department of Physics and Astronomy, University of Kentucky Lexington, KY 40506}
\address{$^{5}$Princeton University, NJ 08544}
\address{$^{6}$Joint Institute for Computational Sciences, Oak Ridge National Laboratory, TN 37831}
\address{$^{7}$Physics Division, Oak Ridge National Laboratory, TN 37831}
\ead{endevee@ornl.gov}

\begin{abstract}
The {\bf t}oolkit for {\bf h}igh-{\bf or}der {\bf n}eutrino-r{\bf ad}iation hydr{\bf o}dynamics (\thornado) is being developed for simulations of core-collapse supernovae (CCSNe) and related problems.  
Current capabilities in \thornado\ include solvers for the Euler equations --- in non-relativistic and special relativistic limits --- and the two-moment model of neutrino transport.  
The spatial discretization in \thornado\ is based on the discontinuous Galerkin (DG) method, which is receiving increased attention from the computational astrophysics community.  
In this paper, we provide an overview of the numerical methods for the Euler equations in \thornado, and present some encouraging preliminary numerical results from a set of basic tests in one and two spatial dimensions.  
\end{abstract}

@@

The {\bf t}oolkit for {\bf h}igh-{\bf or}der {\bf n}eutrino-r{\bf ad}iation hydr{\bf o}dynamics (\thornado) is being developed to simulate neutrino-radiation hydrodynamics in CCSNe and related applications in nuclear astrophysics.  
The spatial discretization of solvers for hyperbolic partial differential equations in \thornado\ is based on the DG method, while we foresee using a combination of spectral and continuous finite element methods for solving elliptic equations; e.g., for Newtonian gravity or relativistic gravity employing the conformal flatness approximation \cite{wilson_etal_1996}.  
With these approaches, we employ high-order discretization techniques based on a common mathematical framework in all the major model components.  
In this paper, we provide an initial description and encouraging demonstration of the DG method implemented in \thornado\ to solve the Euler equations.  
We note that others (e.g., \cite{radiceRezzolla_2011,schaal_etal_2015}) have also investigated the DG method in the context of astrophysical hydrodynamics and reported encouraging results.  
Whether the high-order approach will improve accuracy and efficiency of CCSN models remains to be demonstrated.

\section{Discontinuous Galerkin method}

@@

We let the approximation space for the DG method, $\mathbb{V}^{k}$, be constructed from the tensor product of one-dimensional polynomials of maximal degree $k$.  
Note that functions in $\mathbb{V}^{k}$ can be discontinuous across element interfaces.  
The semi-discrete DG problem is to find $\vect{U}_{h}\in\mathbb{V}^{k}$, which approximates $\vect{U}$ in Eq.~\eqref{eq:extendedEulerCompact}, such that for all $v\in\mathbb{V}^{k}$ and all $\bK\in\mathcal{T}$
\begin{align}
  &\partial_{t}\int_{\bK}\vect{U}_{h}\,v\,dV
  +\sum_{i=1}^{d}\int_{\partial\bK^{i}}\big(\,\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\vect{U}_{h})\,v\big|_{\xH^{i}}-\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\vect{U}_{h})\,v\big|_{\xL^{i}}\,\big)\,d\tilde{\bx}^{i} \nonumber \\
  &\hspace{24pt}
  -\sum_{i=1}^{d}\int_{\bK}\vect{F}^{i}(\vect{U}_{h})\,\pd{v}{i}\,dV
  =\int_{\bK}\vect{S}(\vect{U}_{h})\,v\,dV.  
  \label{eq:semidiscreteDG}
\end{align}
In Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\vect{F}}^{i}(\vect{U}_{h})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
The numerical flux function $\vect{f}^{i}$ is evaluated using values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\vect{F}}^{i}(\vect{U}_{h})=\vect{f}^{i}(\vect{U}_{h}(x^{i,-},\tilde{\bx}^{i}),\vect{U}_{h}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g. in the arguments of $\vect{U}_{h}$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
We use the Harten-Lax-van Leer (HLL) flux \cite{harten_etal_1983} or the HLLC flux \cite{toro_etal_1994,mignoneBodo_2005} for all the numerical experiments presented in Section~\ref{sec:numerical}.

We provide further details on the DG method to arrive at the equations that are actually evolved in \thornado.  
We start by introducing some notation, defining the polynomial expansion for $\vect{U}_{h}$ and the quadrature rules used to evaluate the integrals in Eq.~\eqref{eq:semidiscreteDG}.  
Then we provide explicit expressions for each of the terms in Eq.~\eqref{eq:semidiscreteDG}.  
For simplicity we consider one spatial dimension ($d=1$) and drop indices denoting the spatial dimension.  
In each element $K$, we use a nodal representation of the conserved variables $\vect{U}$; i.e.,
\begin{equation}
  \vect{U}(x,t)\approx
  \vect{U}_{h}(x,t)=\sum_{i=1}^{N}\vect{U}_{i}(t)\,\ell_{i}(x),
  \quad\text{where}\quad
  \ell_{i}(\eta)=
  \prod_{\substack{j=1\\j\ne i}}^{N}\f{\eta-\eta_{j}}{\eta_{i}-\eta_{j}}
  \label{eq:conservedNodalExpansion}
\end{equation}
are Lagrange polynomials defined on the reference element $I = \{ \eta : \eta \in (-0.5,0.5) \}$, and constructed to interpolate the node set $S_{N}=\{\eta_{i}\}_{i=1}^{N}\subset I$.  
The spatial coordinate $x$ and the reference coordinate $\eta$ are related by the mapping $x(\eta)=\xL+(0.5+\eta)\,\dx$.  
Then, for any $\eta_{j}\in S_{N}$, $\ell_{i}(\eta_{j})=\delta_{ij}$, so that $\vect{U}_{h}(x(\eta_{j}),t)=\vect{U}_{j}(t)$.  
We introduce numerical quadratures to evaluate the integrals in Eq.~\eqref{eq:semidiscreteDG}.  
First we define the $M$-point quadrature $Q_{M}:C^{0}(I)\to\mathbb{R}$ with abscissas $\hat{S}_{M}=\{\eta_{q}\}_{q=1}^{M}$ and weights $\{w_{q}\}_{q=1}^{M}$, normalized such that $\sum_{q=1}^{M}w_{q}=1$.  
The $M$-point Legendre-Gauss quadrature, which we use, integrates polynomials of degree $\le 2M-1$ exactly.  
Then, if $P_{h}(x)$ is such a polynomial, we have
\begin{equation}
  \f{1}{\dx}\int_{K}P_{h}(x)\,dx=\int_{I}P_{h}(\eta)\,d\eta=Q_{M}\big[P_{h}\big]\equiv\sum_{q=1}^{M}w_{q}\,P_{h}(\eta_{q}).  
\end{equation}

@@

\section{Preliminary numerical results}
\label{sec:numerical}

In this section we show preliminary numerical results obtained with the DG method implemented in \thornado\ to solve the non-relativistic and special relativistic Euler equations.  
We use the ideal gas equation of state where the pressure $p$ is related to the internal energy density $e$ by $p=(\Gamma-1)e$, and where $\Gamma$ is the (constant) adiabatic index.  
Let the $d$-dimensional computational domain be $D=\times_{i=1}^{d}[\xMin^{i},\xMax^{i}]$, where $\xMin^{i}$ and $\xMax^{i}$ denote the coordinates of the inner and outer boundary in the $i$-th coordinate direction, respectively.  
We use polynomials of degree $k=2$ combined with SSP-RK3 time stepping, and all the tests were run with a Courant-Friedrichs-Lewy (CFL) factor of $C=0.1$.  
We use the HLL Riemann solver when solving the non-relativistic Euler equations and the HLLC Riemann solver when solving the special relativistic Euler equations.  
The purpose of the tests is to gauge the performance of the DG implementation on a set of benchmarks as an initial measure of its suitability for future CCSN simulations.

@@

For the NR Euler equations, the state, flux, and source vectors in Eq.~\eqref{eq:extendedEulerCompact} are
\begin{equation}
  \vect{U}=\big(\rho,\rho u_{j},E\big)^{\trans},~
  \vect{F}^{i}=\big(\rho u^{i},\Pi^{i}_{~j},(E+p)u^{i}\big)^{\trans},~
  \text{and}~
  \vect{S}=\big(0,\f{1}{2}\Pi^{ik}\pd{\gamma_{ik}}{j}-\rho\pd{\Phi}{j},-\rho u^{i}\pd{\Phi}{i}\big)^{\trans},
\end{equation}
where $\rho$ and $u^{i}$ are the mass density and components of the fluid three-velocity, respectively.  
The stress tensor is $\Pi^{i}_{~j}=\rho u^{i} u_{j}+p\delta^{i}_{~j}$, and $E=e+\f{1}{2}\rho u_{i} u^{i}$ is the fluid (internal plus kinetic) energy density.  
The sources are due to curvilinear coordinates and Newtonian gravity.

\subsubsection{Sod shock tube}

@@

\subsubsection{Sedov-Taylor blast wave}

@@

\subsubsection{Shu-Osher shock tube}

@@

\subsubsection{Kelvin-Helmholtz (KH) instability}
This hydrodynamic instability is of significant astrophysical relevance --- including during the explosion phase of CCSNe --- and may occur in the interface separating fluids in relative motion.  
We compute this problem on the unit square $D=[0,1]\times[0,1]$, imposing periodic boundary conditions, and with initial conditions and single-mode perturbation taken from \cite{mcnally_etal_2012}.  
We use $256^{2}$ elements and evolve until $t=3.0$, well into the nonlinear regime.  
Results are displayed in Figure~\ref{fig:KelvinHelmholtz}.  
We have computed two models, one with $C_{\TCI}=0.06$ (left panels) and one without limiting ($C_{\TCI}\to\infty$; right panels).

@@

Early on ($t=1.5$), the results from the two runs are practically indistinguishable, show no sign of developing secondary billows, and agree visually with results presented in \cite{mcnally_etal_2012}.  
When $t=2.4$, the run with $C_{\TCI}=0.03$ has started to develop secondary billows within the coiled-up interface separating dense (white) and less dense (black) fluids, while the model with no limiting has not.  
When $t=3.0$, these secondary instabilities appear to have largely destroyed the coil structure for the model with limiting, while the coils remain intact in the model without limiting (although secondary billows have started to form in this model at this time).  
As discussed in more detail by \cite{mcnally_etal_2012}, the development of the secondary billows may be an artifact of numerical perturbations and diffusion, which is delayed when the resolution is increased.  
Our results are consistent with this in the sense that the less diffusive model develops secondary billows later.

\subsubsection{Liska-Wendroff implosion}

@@

\subsubsection{Standing accretion shock (SAS)}

@@

\subsubsection{Yahil-Lattimer collapse}

The final test of the NR hydrodynamics in \thornado\ involves self-gravity and is due to \cite{yahilLattimer_1982,yahil_1983}.  
It models the self-similar collapse of a polytropic star; i.e., $p=\kappa\rho^{\Gamma}$, where $\kappa$ is the polytropic constant.  
This test is also of immediate relevance to CCSN simulations.  
In \cite{yahilLattimer_1982,yahil_1983}, self-similar solutions to the gravitational collapse problem were constructed for $6/5\le\Gamma<4/3$.   
The solutions smoothly connect an homologously collapsing subsonic inner core (velocity proportional to radius) to a supersonically collapsing outer core in near free-fall ($u\propto r^{-1/2}$).  
\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/YahilLattimerCollapse_MassDensity_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/YahilLattimerCollapse_Velocity_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:YahilLattimer}Results from the Yahil-Lattimer collapse test using $256$ elements.  Results obtained with \thornado\ (dashed black) are compared with the reference solution from Yahil \cite{yahil_1983} (solid red).  The mass density (left panel) and velocity (right panel) are plotted versus radius.  We compare the solutions at select central densities during collapse, approximately $[10^{10},10^{11},10^{12},10^{13},10^{14}]$~g~cm$^{-3}$, which correspond to $(-t)=[51.0,15.0,5.0,1.5,0.5]$~ms.  }
\end{figure}
With two dimensional parameters in the model (the gravitational constant $G$ and the polytropic constant $\kappa$), the dimensionless similarity variable is
\begin{equation}
  X = \kappa^{-1/2} \, G^{(\Gamma-1)/2} \, r \, (-t)^{\Gamma-2},
\end{equation}
where the origin of time is the moment of infinite central density.  
All the hydrodynamic variables can be expressed as a function of $X$, and the time-dependent Euler equations can be recast as a system of ODEs (see \cite{yahil_1983} for details).  
To construct a reference to compare with our numerical results we have solved the ODEs given in \cite{yahil_1983} to obtain these self-similar solutions.

We show results for a model with $\Gamma=1.30$.  
We use spherical polar coordinates, and let $D=[0,1\times10^{5}]$~km, which is covered with $256$ elements.  
The Newtonian gravitational potential is obtained by solving Poisson's equation using a third-order accurate continuous finite element method (e.g., \cite{brennerScott_2008}).  
We use a geometric grid to resolve the mass distribution as the star collapses and the central density increases from about $10^{9}$~g~cm$^{-3}$ to about $10^{14}$~g~cm$^{-3}$.  
The size of the innermost element is set to $1$~km while the last element is about $3\times10^{3}$~km.  
We specify the polytropic constant by setting a reference pressure $p=6\times10^{27}$~erg~cm$^{-3}$ for $\rho=7\times10^{9}$~g~cm$^{-3}$ (reasonable values for a massive star in the pre-collapse stage).  
We also specify the collapse time to $(-t)=150$~ms.  
Results comparing the mass density and velocity of Yahil's reference solution to the solution obtained with \thornado\ using $C_{\TCI}=0.03$ are plotted in Figure~\ref{fig:YahilLattimer}.  
The agreement of the mass density and velocity profiles is good throughout collapse.

@@

\section{Summary and outlook}

We have presented preliminary algorithm details and numerical results for solvers of the non-relativistic and special relativistic Euler equations of gas dynamics as implemented in the toolkit for high-order neutrino-radiation hydrodynamics (\thornado).  
The spatial discretization is based on the DG method and the ODEs resulting from this discretization are integrated in time using SSP-RK methods.  
We employ a spectral-type nodal collocation DG approximation based on Legendre-Gauss points for interpolation and numerical quadrature evaluation \cite{bassi_etal_2013}.  
This choice simplifies expressions for the semi-discretized equations, especially when applied to problems involving curvilinear coordinates encoded in a metric (e.g., in numerical relativity).  
Results from a suite of tests in one and two spatial dimensions --- including problems with strong shocks --- demonstrate reliable performance of the implementation.  
Two tests in spherical symmetry (standing accretion shock \cite{blondin_etal_2003} and Yahil-Lattimer collapse \cite{yahilLattimer_1982,yahil_1983}) demonstrate the DG method's ability to handle conditions relevant to CCSN simulations.  
We will present a more in-depth analysis of the performance of \thornado\ on these (and related) problems in a future study.

@@

\ack Eirik Endeve and Anthony Mezzacappa acknowledge support from the NSF Gravitational Physics Program (NSF-GP 1505933 and 1806692).
